\documentclass[english]{scrartcl}
\usepackage{fontspec}
\usepackage{babel}
\usepackage[backend=biber]{biblatex}
\bibliography{lefschetz_formula.bib}
\usepackage{scrlayer-scrpage}
\usepackage{mathtools, amssymb, amsthm, dsfont}
\usepackage{enumitem}
\setlist[description]{font=\normalfont\bfseries}
\setlist[enumerate]{label=(\arabic*)}
\usepackage{tikz-cd}

\usepackage{hyperref}
\hypersetup{
  pdfauthor=Gesina Schwalbe,
  pdftitle=Lefschetz fixed-point formula
}

% THEOREMS
\theoremstyle{definition}
\newtheorem{Def}{Definition}[section]
\newtheorem{DefLem}[Def]{Definition/Lemma}
\newtheorem{Prop}[Def]{Proposition}
\newtheorem{Thm}[Def]{Theorem}
\newtheorem{Lem}[Def]{Lemma}
\theoremstyle{remark}
\newtheorem{Rem}[Def]{Remark}
\newtheorem{Rev}[Def]{Reminder}

% SHORTENINGS
\newcommand*{\N}{\mathds{N}}
\newcommand*{\Z}{\mathds{Z}}
\newcommand*{\Q}{\mathds{Q}}
\newcommand*{\R}{\mathds{R}}
\newcommand*{\C}{\mathds{C}}
\newcommand*{\fF}{\mathds{F}} % finite field
\newcommand*{\Zmod}[1]{\Z/#1} % Z/nZ
\newcommand*{\Zl}{\Z_l} % l-adic completion of Z
\newcommand*{\Ql}{\Q_l} % quotient field of \Zl
\newcommand*{\F}{\mathcal{F}} % sheaf F
\newcommand*{\G}{\mathcal{G}} % sheaf G
\DeclareMathOperator{\Char}{char} % characteristic of a field
\newcommand*{\End}[1]{\text{End}(#1)} % Endomorphisms
\DeclareMathOperator{\Ext}{Ext} % Ext functor
\DeclareMathOperator{\Tor}{Tor} % Tor functor
\DeclareMathOperator{\Hom}{Hom} % Hom functor
\newcommand*{\id}{\text{id}} % identity
\DeclareMathOperator{\Sh}{Sh} % sheaves over a space
\newcommand*{\idest}{i.e.\ }
\newcommand*{\forexample}{e.g.\ }

\newcommand*{\M}{\Lambda}
% \M_l(r)(R) = {
% for r=0 \ZlZ;
% for r>0 \mu_l(R)^{\otimes r};
% for r<0 Hom_{\Zmod{l}}(\mu_l(R)^{\otimes -r}, \Zmod{l}) } 
\newcommand*{\tr}[1]{\text{tr}\left(#1\right)} % trace isomorphism
\newcommand*{\intProd}[2]{{#1\cdot#2}} % intersec. product in the Chow ring
\newcommand*{\intNum}[1]{{\left\langle{#1}\right\rangle}} % intersection number (= tr(intersection product))
\newcommand*{\Graph}[1]{{\Gamma_{#1}}} % graph of a map
\newcommand*{\Diag}[1]{{\Delta_{#1}}} % diagonal (both map and subscheme)
\newcommand*{\trace}[2]{{\text{trace}\left(#1 \,\middle|\, #2 \right)}} % trace of map (#1) in vector space (#2)
\newcommand*{\cl}[2]{{\text{cl}_{#1}\left(#2\right)}} % cycle class map
\newcommand*{\one}[1]{{1_{#1}}}%{H^*(#1)}}
\newcommand*{\dual}[2]{{#1_{#2}^*}} % dual basis vector
\let\altphi\phi
\renewcommand*{\phi}{\varphi}
\newcommand*{\Poincare}{{\text{Poincaré \ref{poincare}}}}
\newcommand*{\Hc}{{H_c}} % cohomology with compact support

\begin{document}
% HEAD/FOOT
\clearpairofpagestyles
\ohead{Universität Regensburg}
\ihead{Gesina Schwalbe}
\cfoot*{\pagemark}

% TITLE
\title{Poincaré Duality and\\
  Lefschetz Fixed-Foint Formula}
\subject{Seminar:
  Deligne's proof of the Weil Conjecture%\\by Federico Binda, Prof. Dr. Moritz Kerz
}
%\subtitle{Lecture Notes}
\author{Gesina Schwalbe}
%\date{}
\maketitle
\tableofcontents

\section{Motivation}
From topology the following theorem is known that relates the fixed
points of a morphism (counted with multiplicity) with homology:
\begin{Thm}[Lefschetz-Hopf]
  Let $X$ be a compact, smooth manifold and
  $\phi\colon X\to X$ a morphism. Then
  \begin{gather*}
    \sum_{x\in\operatorname{Fix}(\phi)} \operatorname{i}(\phi,x)
    =\sum_{r\in\N}(-1)^r \trace{\phi^*}{H_\text{sing}^r(X,\Q)}\;,
  \end{gather*}
  where $\operatorname{Fix}(\phi)$ are the fixed points of $\phi$ and
  $\mathrm i(\phi,x)$ is the multiplicity of a fixed point $x$ of $\phi$
  in $X$.
\end{Thm}
The proof uses the Poincaré duality theorem which says
\begin{Thm}[Poincaré]
  For a smooth manifold $X$ holds for all $r\in\Z$
  \begin{gather*}
    H_{\text{c,sing}}^r(X,\Z) \cong H_{\dim(X)-r}^{\text{sing}}(X,\Z)
  \end{gather*}
  where $H_{\text{c,sing}}(X,\Z)\coloneqq
  \lim_{Z\subset X \text{cpt.}}H_{Z,\text{sing}}^r(X,\Z)$
  is cohomology with compact support for $X$
  (write $H_{Z,\text{sing}}^r(X,\Z)$ for $r$th relative
  $\Z$-cohomology of the pair $(X,Z)$).
\end{Thm}
The aim of this lecture is to introduce an analogue statement for
schemes instead of topological spaces as well as give a proof and an
overview of the necessary preliminaries.
Some convenient correspondences are
\begin{itemize}
\item an algebraically closed field $k$ instead of $\R$ as ground field,
\item $k$-\emph{variety}
  (\idest separated, geometrically reduced scheme of finite type over $k$)
  instead of \emph{$\R$-manifold}
\item \emph{complete} instead of \emph{compact},
\item \emph{non-singular} instead of \emph{smooth}
\item \emph{morphism of varieties} instead of \emph{morphism of manifolds}
\item \emph{étale cohomology} instead of \emph{singular homology}
  (mind that they coincide for non-singular varieties over $\C$ and
  finite coefficient group \cite[see][Thm.~21.1]{milne})
\item intersection number $\intNum{\intProd{\Graph{\phi}}{\Diag{X}}}$
  (also counting fixed points of $\phi$ with multiplicity)
  instead of $\sum_{x\in\operatorname{Fix}(\phi)}\mathrm{i}(\phi,\cdot)$.
\end{itemize}
We also need to consider modified coefficients since étale cohomology
does not behave well for non-finite coefficient groups. However,
passing to limits we can consider for $X$ (étale) scheme the étale
cohomology with coefficients in $\Ql$ as follows
\begin{gather*}
  H^r(X,\Ql)\coloneqq
  \left(\lim_{n\to\infty}H^r(X,\Zmod{l^n})\right)\otimes_{\Zl}\Ql
\end{gather*}
Thus the final statement of the Lefschetz fixed point formula shall be
\begin{Thm}[Lefschetz]\label{lefschetzthm:motivation}
  Let $k$ be an algebraically closed field, $X$ a non-singular,
  complete $k$-variety, and let $\phi\colon X\to X$ be an
  endomorphism. Then
  \begin{gather*}
    \intNum{\intProd{\Graph{\phi}}{\Diag{X}}}
    = \sum_{r\in\N}(-1)^r\trace{\phi^*}{H^r(X,\Ql)}
  \end{gather*}
\end{Thm}

We instantly get the following two applications:
\begin{description}
\item[Zeta function] For
  $k=\fF_{q}$ a finite field ($q\in\N$),
  $X$ a non-singular variety over $k$ and
  $\phi=F^n$ the $n$th Frobenius homomorphism
  we get
  \begin{gather*}
    \#X(\fF_{q^n}) = \intNum{\intProd{\Graph\phi}{\Diag X}}
  \end{gather*}
  which is essential to further investigate the Zeta function of $X$.
\item[Euler characteristic] Taking $k$ and $X$ as in the Theorem and
  $\phi=\Delta$ we see that the étale Euler characteristic of $X$ is
  just its self-intersection number, \idest
  \begin{gather*}
    \intNum{\intProd{\Delta}{\Delta}} = \chi_{\text{ét}}(X)
  \end{gather*}
\end{description}

For the proof of \ref{lefschetzthm:motivation}
some further nice properties of étale cohomology for
(nice) varieties will be needed, such as:
a Künneth isomorphism (\autoref{kuennethiso}),
a cycle class map that gives a connection to the Chow ring and
intersection product (\autoref{cycleclassmap}),
Poincaré duality with a trace map (\autoref{poincare}),
as well as pushforwards on cohomology (\autoref{gysinmaps}).
To establish these, one needs for étale cohomology the notion of
cohomology with compact support (\autoref{compactcoh}),
relative cohomology (\autoref{relcoh}),
long exact sequences of (compact) pairs (\autoref{LES}),
Gysin sequences (\autoref{gysinsequ}),
and most of all a cup product, \idest a graded ring structure
(\autoref{cupproduct}).
All of this will be introduced in the following.

\section{Preliminaries on étale Cohomology}
It will be assumed the reader knows about the definition and basic
properties of étale cohomology. If not stated otherwise, for a
(Zariski) scheme $X$ and a (Zariski) sheaf of modules $\F$ over $X$ we
denote with $H^r(X,\F)$ the $r$th étale cohomology group of the étale
sheaf $\F$ over the étale scheme of $X$
\cite[compare][Chap.~6, Examples of Sheaves on $X_\text{ét}$]{milne}.

Also recall that for $\F$ torsion sheaf or $X$ complete these groups
are finite \cite[\forexample][Thm.~19.1]{milne}.

\subsection{Étale Cohomology with compact support}
In this section let $k$ be an algebraically closed field,
$U$ a variety over $k$,
and $\F$ be a torsion sheaf over $U$.

The definition of cohomology with compact support as
$\left(\lim_{Z\subset X \text{cpt.}}H_Z^r(X,\Z)\right)$
cannot be copied from topology but has to be adjusted.
However, we will see later (namely \autoref{rem:cptvsrelcohomology}),
that at least in degree 0 the following definition
and the expression above coincide.

First recall the following
\begin{Rev}[Nagata compactification]\label{nagata}
  There always exists a complete $k$-variety $X$ and an open, dense
  immersion $j\colon U\to X$. $j$ is also called a
  \emph{compactification} of $U$.
\end{Rev}
\begin{Rev}[Extension by zero]\label{def:extbyzero}
  For an open immersion $j\colon U\to X$ into
  a scheme $X$, $j_!$ (called the \emph{extension by zero} of $j$)
  is defined as
  \begin{gather*}
    j_!\colon \Sh(U)\to \Sh(X)\;,\qquad
    j_!\F \coloneqq \text{sheafification of}\quad
    V\mapsto\begin{cases}
      \F(V) & V\subset U\\
      0     & \text{else}
    \end{cases}
  \end{gather*}
\end{Rev}

\begin{DefLem}\label{def:cptcohomology}
  Let $j\colon U\to X$ be a compactification of $U$. Then the
  cohomology with compact support in degree $r\in\Z$ is defined as
  \begin{gather*}
    \Hc^r(U,\F) \coloneqq H^r(X,j_!(\F))
  \end{gather*}
  This definition is independent of the choice of compactification.
  \begin{proof}[proof (scetch)]
    The set of compactifications of $U$ is partially ordered by proper
    morphisms:
    For two compactifications $j_1\colon U\to X_1$,
    $j_2\colon U\to X_2$ the closure
    $X\coloneqq\overline{(j_1,j_2)(U)}$ in $X_1\times_X X_2$ is again
    a compactification and the projections on $X_1$, $X_2$ are proper.
    Thus one only needs to check $\Hc^r(X,j_!\F)\cong\Hc^r(X',j_!'\F)$
    for compactifications $j$, $j'$ and a proper map $\pi$ such that
    the following diagramm commutes
    \begin{center}
      \begin{tikzcd}
        U\arrow[r,"j"]\arrow[dr,"j'"{left}, bend right=20]
        & X\arrow[d,"\pi"]\\
        & X'
      \end{tikzcd}
    \end{center}
    This then can be shown using a version of the Leray spectral
    sequence and the proper base change theorem.
  \end{proof}
\end{DefLem}

\begin{Rem}
Note that this is no derived functor!
However, since $j_!$ is exact for a compactification $j$ of $U$, we
still get a long exact sequence of cohomology out of a short exact
sequence of a short exact sequence of sheaves on $U$
(just use the definition and the long exact sequence of cohomology for
$H^*(X,\cdot)$).
\end{Rem}

\subsection{Ext-Pairings, Cup Product and Künneth formula}
Again let $k$ be an algebraically closed field throughout the
section.

The foundation of all further investigations will be the graded
ring structure on the sum of the cohomology groups in all degrees
(with appropriate coefficients). The product will be a cup product
denoted $\cup$ to which we here give an approach via pairings of
Ext-groups. 
However, there are concrete descriptions using Čech-cohomology
\cite[see][Chap.~10 and Chap.~22, Cup-products]{milne}
or using the Künneth isomorphism defined later on (see
\autoref{kuennethiso}, \autoref{rem:cupwithkuennethiso}).

\begin{Rem}[Yoneda product]\label{yonedaproduct}
  For an abelian category $\mathcal{A}$ with enough injectives and
  any objects $A,B,C$ of $\mathcal{A}$, $r,s\in\N$ there exists a pairing
  \begin{align*}
    \Ext^r(A,B)\otimes\Ext^s(B,C) &\overset{\cup}{\longrightarrow}
    \Ext^{r+s}(A,C)
    % \intertext{which is induced by}
    \\
    \text{induced by}\qquad
    \Hom(A,B)\otimes\Hom(B,C) &\longrightarrow \Hom(A,C)\;,\quad
                                f\otimes g\mapsto f\circ g
  \end{align*}
  \begin{proof}[idea for construction]
    The elements of an $r$th Ext-group can be described as extensions
    (\idest sequences) of length $r$.
    Define $\cup$ to be the concatenation of two such sequences. 
  \end{proof}
\end{Rem}

\begin{Prop}[Cup product]\label{def:cupproduct}
  For a $k$-variety $X$, and $\F,\G\in\Sh(X)$ get for all $r,s\in\Z$
  natural pairings
  \begin{gather*}
    H^r(X,\G)\otimes H^r(X,\F) \overset{\cup}\longrightarrow
    H^{r+s}(X,\F\otimes\G)
  \end{gather*}
  This is associative as well as graded commutative, \idest $x\cup y =
  (-1)^{rs}y\cup x$ for $x\in H^r(X,\F)$, $y\in H^s(X,\G$
  \cite[][Rem. 1.18]{milnebook}.
  \begin{proof}[idea for construction]
    The cup product is the map making the following diagram commute
    where the lower row is the \nameref{yonedaproduct}
    \begin{center}
      \begin{tikzcd}[column sep=tiny]
        H^r(X,\F) \arrow[r, phantom, "\otimes"{description}] \arrow[d, equals]&
        H^r(X,\G) \arrow[r] \arrow[d]
        & H^{r+s}(X,\F\otimes\G) \arrow[d,equals]\\
        \Ext_X^r(\mathcal O_X,\F) \arrow[r, phantom, "\otimes"{description}]&
        \Ext_X^s(\F,\F\otimes\G) \arrow[r]
        & \Ext_X^{r+s}(\mathcal O_X,\F\otimes\G)
      \end{tikzcd}
    \end{center}
    This already implies associativity (compare the construction of the
    Yoneda product in \ref{yonedaproduct}).
    
    There is also a more general approach to define cup products for
    derived functors. This states, that for any right exact functor
    $\Gamma\colon \Sh(X)\to\text{AbGrp}$ fulfilling certain nice
    properties, a bifunctor
    $\Gamma(\F)\otimes\Gamma(\G)\to\Gamma(\F\otimes\G)$
    extends to a cup product pairing
    \begin{gather*}
      \operatorname{R}^r\Gamma(\F)\otimes
      \operatorname{R}^s\Gamma(\G) \to
      \operatorname{R}^{r+s}\Gamma(\F\otimes\G)
    \end{gather*}
    \cite[see][Prop. V.1.16]{milnebook}
  \end{proof}
\end{Prop}

We can now proceed with the definition of the graded ring of
cohomology groups by choosing appropriate coefficients.
\begin{Def}\label{def:coefficients}
  Let $X$ be a $k$-variety,
  $n\in\N$ such that $\Char(k)\nmid n$,
  and $r\in\Z$.
  \begin{enumerate}
  \item Let $\Lambda\coloneqq\Lambda_n$ be the constant sheaf of
    $\Zmod{n}$ over $X$.
  \item For an integral domain $R$ let
    $\mu_n(R)\coloneqq\{ u\in R^\times\;|\;u^n=1_R\}$ be the $n$th
    roots of unity in $R$. Define
    \begin{align*}
      \SwapAboveDisplaySkip
      \mu_n(R)^{\otimes r} &\coloneqq \begin{cases}
        \mu_n(R)^{\otimes r} &r>0\\
        \Zmod{n} &r=0\\
        \Hom_{\Zmod{n}}(\mu_n(R)^{\otimes (-r)},\Zmod{n}) &r<0
      \end{cases}
    \end{align*}
    Note that if $R$ contains all $n$th roots of unity, then a
    primitive $n$th root of unity determines isomorphisms
    $\mu_n(R)^{\otimes s}\cong\Zmod{n}$ for all $s\in\Z$.
  \item We define the $r$th twist of $\Lambda$ as
    \begin{gather*}
      \Lambda(r) \coloneqq \text{ sheafification of }\quad
      U\mapsto \mu_n(\Gamma(U,\mathcal O_U)^{\otimes r}
    \end{gather*}
    As explained above, for $k$ algebraically closed $\Lambda(r)$ is
    (non-canonically) isomorphic to $\Lambda$.
    However, choosing a primitive $n$th root of unity determines
    isomorphisms $\Lambda\cong\Lambda(s)$ for all $s\in\Z$ and
    $\Lambda(r)^\vee\coloneqq \Hom_\Lambda(\Lambda(r),\Lambda)
    \cong\Lambda(-r)$.
  \item For an $n$-torsion sheaf over $X$ (\idest a $\Lambda$-module)
    set
    \begin{align*}
      \F(r) &\coloneqq \F\otimes_\Lambda \Lambda(r)\\
      \F(r)^\vee &\coloneqq \Hom_\Lambda(\F,\Lambda(r))\cong \F(-r)
    \end{align*}
  \end{enumerate}
\end{Def}

One may notice an analogue of twisting sheaves in sheaf cohomology,
where $\mathcal O_X$-modules are considered.

With the coefficients defined above we can now state the ring
structure
as follows:
\begin{DefLem}
  For a $k$-variety $X$, and $n\in\Z$ such that $\Char(k)\nmid n$
  define for $r\in\Z$
  \begin{align*}
    H^r(X) \coloneqq H^r(X,\Lambda(r))\;,\qquad
    H^*(X) \coloneqq \bigoplus_{r\in\N}H^r(X)
  \end{align*}
  \begin{enumerate}
  \item $H^*(X)$ together with the cup product as in \ref{def:cupproduct}
    $H^*(X)\otimes H^*(X)\xrightarrow{\cup} H^*(X)$
    is a graded commutative ring
    (follows directly from the properties of the cup product).
  \item The canonical isomorphism $\Zmod{n}\cong H^0(X)$ makes
    $H^*(X)$ an $\Zmod{n}$-Algebra with unit $1_X$,
    which is the image of $1_{\Zmod{n}}$ under the isomorphism.
  \item For any morphism of $k$-varieties $\phi\colon X\to Y$ the
    induced map $\phi^*\colon H^*(Y)\to H^*(X)$ is a homomorphism of
    graded rings
    (follows from the naturality of the cup product).
  \end{enumerate}
\end{DefLem}

A first heavily simplifying observation is the Künneth formula which
produces the cohomology ring of a product out of the cohomology rings
of each factor.

\begin{Thm}[Künneth formula]\label{kuennethiso}
  Let $X$ and $Y$ be $k$-varieties
  ($k$ only needs to be separably closed with $\Char(k)\nmid n$).
  Further let $\F$ and $\G$ be constructible, flat $\Lambda_n$-modules
  over $X$ resp. $Y$.
  Consider the projection maps
  $p\colon X\times_k Y\to X$ and
  $q\colon X\times_k Y\to Y$ and denote
  $\F\boxtimes\G\coloneqq p^*\F\otimes q^*\G$. 
  \begin{enumerate}
  \item   Then for any $m\in\N$ there is an exact sequence
    \begin{center}
      \begin{tikzcd}[row sep=tiny]
        0 \rar
        &\displaystyle\bigoplus_{\mathclap{r+s=m}}H^r(X,\F)\otimes H^s(Y,\G) \rar
        &H^m(X\times_k Y,\F\boxtimes\G) \dar\\
        &&\displaystyle\bigoplus_{\mathclap{r+s=m+1}}\Tor_1^{\Zmod{n}}\left(
          H^r(X,\F), H^s(Y,\G)
        \right) \rar
        &0
      \end{tikzcd}
    \end{center}
    This also holds if one exchanges étale cohomology with étale
    cohomology with compact support.
    The second map is defined by
    \begin{center}
      \begin{tikzcd}[row sep=small]
        \displaystyle\bigoplus_{\mathclap{r+s=m}}
        H^r(X,\F)\otimes H^s(Y,\G)
        \dar&
        \displaystyle\sum_{\mathclap{i}}a_i\otimes b_i
        \arrow[dd,mapsto]\\
        \displaystyle\bigoplus_{\mathclap{r+s=m}}
        H^r(X\times_k Y,q^*\F)\otimes H^s(X\times_k Y,q^*\G)
        \arrow[d,"\cup"]\\
        H^m(X\times_k Y,\F\boxtimes\G)
        &\displaystyle\sum_{\mathclap{i}}p^*a_i \cup q^*b_i
      \end{tikzcd}
    \end{center}
  \item For $X$ and $Y$ complete the above map is an isomorphism. 
  \end{enumerate}
  \begin{proof}Skipped.\end{proof}
\end{Thm}

\begin{Rem}
  For $X$ and $Y$ complete $k$-varieties we obtain a natural
  isomorphism of graded rings, the so-called Künneth isomorphism:
  \begin{align*}
    H^*(X)\otimes H^*(Y) &\rightarrow H^*(X\times_k Y)\\
    a\otimes b &\mapsto p^*(a) \cup q^*(b)
  \end{align*}
  Mind that functoriality makes it compatible with the limit maps
  $\Lambda_{l^{n+1}}(r)\to\Lambda_{l^n}(r)$.
\end{Rem}

\subsection{The Chow Ring and the Cycle Map}

\section{Poincaré Duality}
\subsection{Trace Map}
\subsection{Poincaré Duality}
\begin{Thm}[Poincaré]
  
\end{Thm}
\subsection{Gysin Maps}

\section{Lefschetz Fixed-Point Formula}

\begin{Thm}[Lefschetz fixed-point formula]\label{lefschetzthm}
  Let $X$ be a non-singular, complete variety over an algebraicly
  closed field $k$.
  Let further $\phi\colon X\to X$ be a regular Endomorphism.
  Then
  \begin{gather*}
    \intNum{\intProd{\Graph{\phi}}{\Diag{X}}}
    = \sum_{r\in\N} (-1)^r \trace{\phi^*}{H^r(X,\Ql)}
  \end{gather*}
  where $\trace{\phi^*}{H^r(X,\Ql)}$ is the trace of $\phi^*$ as
  Endomorphism of a $\Ql$-vector space. 
  The sum is finite due to the vanishing
  of cohomology in degree greater than $2\dim(X)$. 
\end{Thm}
Note that $\intNum{\intProd{\Graph{\phi}}{\Diag{X}}}$ is exactly the number of
fixed points of $\phi$, hence the name of this formula.


\subsection{Proof}
\begin{Rem}
  In the following we will work with
  $H^*(X)\coloneqq\sum_{r\in\N} H^r(X, \Ql(r))$, where
  \begin{gather*}
    H^r(X) = H^r(X,\Ql(r)) \coloneqq
    \left(\lim_{n\to\infty} H^r(X,\M_{l^n}(r))\right) \otimes \Ql
  \end{gather*}
  All the maps and identities we have seen so far are compatible with
  taking the limit and tensoring with $\Ql$, namely
  cup product,
  trace isomorphism,
  Poincaré duality,
  cycle class map and hence
  pushforward and pullback on cohomology as well as
  the Künneth isomorphism.
  Thus $H^*(X)$ together with the induced cup product is a graded ring
  fulfilling these properties.
  For $X$, $\phi$ as above we will deduce the formula
  \begin{gather*}
    \intNum{\intProd{\Graph{\phi}}{\Diag{X}}}
    = \sum_{r\in\N} (-1)^r \trace{\phi^*}{H^r(X)}
  \end{gather*}
  This already implies the theorem
  (\idest $\trace{\phi^*}{H^r(X)})=\trace{\phi^*}{H^r(X,\Ql)}$),
  because:
  Since $k$ is algebraicly closed---especially it contains all
  roots of unity---we can choose for $r=1$ (hence inductively all
  $r\in\N$) and $n\in\N$ an isomorphism
  \begin{gather*}
    \M_{l^n}(r)\coloneqq\mu_{l^n}^{\otimes_k r} \cong \Zmod{(l^n)}
  \end{gather*}
  So $H^r(X)\cong H^r(X,\Ql)$ as $\Ql$-vector spaces.
  One now has to observe, that the traces of $\phi^*$ commute with
  this isomorphism.
  % TODO!!
\end{Rem} 

\begin{Rev}
  Let $V$ be a finite dimensional $K$-vector space, $(e_i)_{i=1}^n$ a basis,
  and $(\check e_i)_{i=1}^n$ the corresponding dual basis.
  Denote the evaluation by
  \begin{align*}
    \cup\colon V\otimes V^* \to K,\quad
    v\otimes \omega \mapsto v\cup\omega\coloneqq \omega(v)
  \end{align*}
  Then the trace of some $f\in\End(V)$ is defined as
  \begin{gather*}
    \trace{f}{V} \coloneqq \sum_{i=1}^n f(e_i)\cup \check e_i
  \end{gather*}
\end{Rev}

\begin{proof}[proof of \autoref{lefschetzthm}]~
  \begin{enumerate}[label={Step \arabic*.}]
  \item %[25.3]
    First we need the following identity:
    \begin{Lem}\label{step1}
      For a regular morphism $\phi\colon X\to Y$ of complete varieties
      and $y\in H^*(Y)$ holds
      \begin{gather*}
        p_*\left(\cl{X\times Y}{\Graph{\phi}} \cup q^*(y)\right)
        = \phi^*(y)           
      \end{gather*}
      \begin{proof}
        Since $Y$ is separated, $\Graph{\phi}\colon X\to X\times_k Y$
        is a closed immersion, hence proper, and we can apply the 
        \nameref{projectionformula}. % REF MISSING
        Furthermore, by functoriality of the cycle class map we have
        \begin{gather}
          \cl{X\times_k Y}{\Graph{\phi}} \coloneqq
          \cl{X\times_k Y}{(\Graph{\phi})_*(X)} =
          (\Graph{\phi})_*(\cl{X}{X}) \overset{\ref{ex:clxx}}{=} % REF MISSING          
          (\Graph{\phi})_*\left(\one{X}\right)
        \end{gather}
        Altogether we get
        \begin{align*}
          p_*\left(\cl{X\times Y}{\Graph{\phi}} \cup q^*(y)\right)
          &=  p_*\left(
            (\Graph{\phi})_*\left(\one{X}\right) \cup q^*(y)
            \right)\\
          &\overset{\mathllap{\nameref{pushforward:projectionformula}}}{=}
            p_*(\Graph{\phi})_*\left( \one{X} \cup (\Graph{\phi})^*q^*(y) \right)\\
          &\overset{\mathllap{\eqref{pushforward:composition}}}=
            (p\circ\Graph{\phi})_* \left(
            \one{X} \cup (q\circ\Graph{\phi})^*(y) \right)\\
          &\overset{\mathllap{\text{Def. $\Graph\phi$}}}=
            \id_* \left( \one{X} \cup \phi^*(y) \right)\\
          &= \phi^*(y)
        \end{align*}
      \end{proof}
    \end{Lem}
  \item Now we can explicitly describe
    $\cl{X\times_k X}{\Graph{\phi}}$:
    \begin{Lem}
      Let $X$ be a complete and non-singular variety of dimension $d$
      and $\altphi\in\End(X)$ regular.
      Write $e^{2d}\in H^{2d}(X)$ for the canonical generator
      with $\tr{e^{2d}}=1\in\Ql$.
      Furthermore, let $(b_k)_k$ be a basis of $H^*(X)$ as
      $\Ql$-vector space and $(f_k)_k$ be the dual Basis with respect
      to the cup product, \idest $b_k\cup f_l=\delta_{kl}e^{2d}$.
      % ?? HOW IS THIS CONSTRUCTED? / WHY DOES THIS EXIST?
      Then
      \begin{gather*}
        \cl{X\times_k X}{\Graph{\phi}} =
        \sum_{k}^{2d} \phi^*(b_k)\otimes  f_k
      \end{gather*}
      \begin{proof}
        By the \nameref{künnethiso} % REF MISSING
        $H^*(X\otimes X) \cong H^*(X)\otimes_{\Ql}H^*(X)$ is a
        $H^*(X)$-module with basis
        $(\one{X}\otimes f_k)_{k}$.
        Therefore, there are unique factors $a_{k}$ in $H^*(X)$ s.t.
        \begin{gather}
          \label{eq:step2}
        \cl{X\times_k Y}{\Graph{\phi}} =
        \sum_{k} a_k\otimes f_k
      \end{gather}
      Explicitly
      \begin{align*}
        a_j
        = p_*(a_j\otimes e^{2d})
        &= p_*\Big(
          \sum_k(a_k\otimes f_k)\cup (\one{X}\otimes b_j)
        \Big)\\
        &\overset{\mathllap{\eqref{eq:step2}}}= p_*\left(
          \cl{X\times_k X}{\Graph{\phi}} \cup q^*(b_j)
          \right)\\
        &\overset{\mathllap{\ref{step1}}}= \phi^*(b_j) 
      \end{align*}
      \end{proof}
    \end{Lem}
  \item
  For all $r\in\{0,\dotsc,2d\}$ let
  \begin{itemize}
  \item $(e_i^r)_{i}$ be a basis of $H^r(X)$ as $\Ql$-vector space,
  \item $(\check e_i^{2d-r})_{i}$ be the dual basis of $H^{2d-r}(X)\cong (H^{r}(X))^\vee$,
    \idest $e^r_i\cup \check e_j^{2d-r}= \delta_{ij}e^{2d}$.
  \end{itemize}
  Note that then $e^r_i\cup \check e_j^{2d-s}=\delta_{ij}\delta_{rs}e^{2d}$,
  % ?? WHY?
  \idest they fulfill the properties of the Lemma
  and we get for $\phi$ and $\id$
  \begin{align*}
    \cl{X\times_k X}{\Graph{\phi}}
    &= \sum_{i,r} \phi^*(e_i^r)\otimes \check e_i^{2d-r}\\
    \cl{X\times_k X}{\Diag{X}}
    &= \sum_{i,r} e_i^r\otimes \check e_i^{2d-r} =
      \sum_{i,r} (-1)^{r\cdot (2d-r)}\check e_i^r\otimes e_i^{2d-r} =
      \sum_{i,r} (-1)^{r}\check e_i^{2d-r}\otimes e_i^r\\
    \cl{X\times_k X}{\intProd{\Graph{\phi}}{\Diag{X}}}
    &= \cl{X\times_k X}{\Diag{X}} \cup \cl{X\times_k X}{\Graph{\phi}}\\
    &= \sum_{i,r,j,s} (-1)^{r}
      \left(\phi^*(e_i^r)\cup \check e_j^{2d-s}\right)
      \otimes \left(\check e_i^{2d-r}\cup e_j^s\right)\\
    &= \sum_{i,r} (-1)^{r}
      \left(\phi^*(e_i^r)\cup \check e_i^{2d-r}\right)
      \otimes e^{2d}\\
    &\overset{\mathllap{\Poincare}}=
      \sum_{r} (-1)^r \trace{\phi^*}{H^r(X)} e^{2d} \otimes e^{2d}
  \end{align*}
  Altogether, using
  \begin{align*}
    \tr{e^{2d} \otimes e^{2d}}
    % TODO!! ??
    % &= \tr{p^*(e^{2d})\cup q^*(e^{2d})}\\
    % &= \tr{p^*(e^{2d})}\cdot\tr{ q^*(e^{2d})}\\
    % &\overset{\eqref{pushforward:trace}=
    %   \tr{e^{2d}}\cdot\tr{e^{2d}}\\
    %   \overset{\text{Def.}}= 1\cdot 1
    = \tr{\Diag{X}_*(e^{2d})}
    \overset{\eqref{pushforward:trace}}= \tr{e^{2d}}
    \overset{\text{Def.}}= 1\in\Ql
  \end{align*}
  we get our final result:
  \begin{gather*}
    \intNum{\intProd{\Graph\phi}{\Diag X}} \coloneqq
    \tr{\cl{X\times_kX}{\intProd{\Graph\phi}{\Diag X}}}
    = \sum_{r\in\N} (-1)^r \trace{\phi^*}{H^r(X,\Ql)}
  \end{gather*}
\end{enumerate}
\end{proof}

\subsection{Other Cases}

\nocite{*}
\printbibliography
\end{document}
